The noise-marginalized optimal statistic: A robust hybrid frequentist-Bayesian statistic for the stochastic gravitational-wave background in pulsar timing arrays

### 2018-05-17 manuscript/manuscript.tex
fixed wording, etc

\begin{abstract}
Observations have revealed that nearly all galaxies 
contain supermassive black holes (SMBHs) at their centers. 
When galaxies merge, these SMBHs form SMBH binaries (SMBHBs), 
which emit low-frequency gravitational waves (GWs). 
The incoherent superposition of these sources produce a stochastic GW background (GWB) 
that can be observed by pulsar timing arrays (PTAs). 
The optimal statistic is a frequentist estimator of the amplitude of the GWB 
that specifically looks for the spatial correlations between pulsars induced by the GWB. 
In this paper, we introduce an improved method for computing the optimal statistic 
that marginalizes over the red noise in individual pulsars. 
We use simulations to demonstrate that this method 
more accurately determines the strength of the GWB, 
and we use the noise-marginalized optimal statistic 
to compare the significance of monopole, dipole, and Hellings-Downs spatial correlations 
and perform sky scrambles.
\end{abstract}

@@

However, when pulsars have significant red noise, 
the optimal statistic gives biased results 
due to the strong covariance 
between the individual red noise parameters and the GWB amplitude. 
Many individual pulsars show evidence for red noise \citep{lam+2017,abb+17}, 
and uncertainty in the position of the Solar System barycenter (SSB) leads to 
a common red process in all pulsars \citep{abb+17b}. 
%In this paper, we present a technique for improving the accuracy of the optimal statistic 
%by marginalizing the optimal statistic over the individual red noise parameters 
%using the posterior distributions from a full Bayesian analysis. 
%This hybrid approach produces a more accurate 
%estimate of \Agw\ and its uncertainty. 
Here we present a technique for improving the accuracy of the optimal statistic by including an additional step: 
marginalizing over the individual pulsar red noise parameters 
using the posterior distributions from a full Bayesian analysis of all the pulsars. 
This hybrid approach produces a more precise estimate of the GWB amplitude \Agw\ 
and its uncertainty, while requiring only a few minutes more than the more traditional method of computation. 
%This method is more computationally intensive than previous methods 
%for computing the optimal statistic, but still requires only minutes. 
%Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
%can be used to compute the optimal statistic with any choice of the ORF, 
%allowing additional spatial correlations to be tested very quickly.
Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
can be used to compute the optimal statistic for any choice of spatial correlations 
simply by changing the overlap-reduction function (ORF). 
For example, clock errors lead to a common red signal with monopole spatial correlations \citep{hcm+2012}, 
while uncertainty in the SSB produces dipole spatial correlations \citep{chm+2010}. 
This technique is used to perform the frequentist searches 
for common red signals with Hellings-Downs, 
monopole, and dipole spatial correlations 
in the NANOGrav 11-year data set \citep{abb+17b}.

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic. 
We use simulations to compare the noise-marginalized optimal statistic 
to the optimal statistic computed with fixed noise. 
In Sec.~\ref{sec:spatial} we determine how well 
the noise-marginalized optimal statistic can 
differentiate between monopole, dipole, and Hellings-Downs spatial correlations. 
In Sec.~\ref{sec:skyscrambles} we use the noise-marginalized optimal statistic 
to perform sky scrambles, 
which assess the significance of Hellings-Downs spatial correlations 
by scrambling the pulsars' sky positions \citep{cs2016, tlb+2017}. 
We summarize our results in Sec.~\ref{sec:conclusion} 
as well as discuss future applications of the noise-marginalized optimal statistic.

@@

The optimal statistic is constructed from the autocovariance and cross-covariance matrices 
$C_a$ and $S_{ab}$, 
\begin{eqnarray}
	C_a &=& \left\langle \res_a \res_a^T \right\rangle \,, \\
	S_{ab} &=& \left. \left\langle \res_a \res_b^T \right\rangle \right|_{a \neq b} \,,
\end{eqnarray}
where $\res_a$ is a vector of the residuals of the $a$th pulsar in the PTA. 
For the GWB with power spectral density (PSD) $P_\mathrm{gw}(f)$ 
and overlap reduction function (ORF) $\Gamma_{ab}$, 
the cross-covariance matrices are
\begin{equation}
	S_{ab} = F_a \, \phi_{ab}^\mathrm{gw} \, F_b^T \,,
\end{equation}
where
\begin{equation}
	\phi_{ab}^\mathrm{gw} = \Gamma_{ab} \, P_\mathrm{gw}(f) \,,
	\label{eq:phi_gw}
\end{equation}
and the ORF is the Hellings-Downs curve \citep{hd1983}, 
\begin{eqnarray}
	\Gamma_{ab} &=& \frac{1}{2} \left[ 1 - \frac{1}{2} \left( \frac{1-\cos\theta_{ab}}{2} \right) \right. \nonumber \\
		&& \left. + 3 \left( \frac{1-\cos\theta_{ab}}{2} \right) \ln \left( \frac{1-\cos\theta_{ab}}{2} \right) \right]  \,,
\end{eqnarray}
where $\theta_{ab}$ is the angle between the pulsars. 
We model the PSD of the GWB as a power law:
\begin{equation}
	P_\mathrm{gw}(f) = \frac{\Agw^2}{12\pi^2} \left( \frac{f}{f_\mathrm{yr}} \right)^{-\gamma} \,,
\end{equation}
where $\gamma = 13/3$ assuming SMBHBs evolve solely due to GW emission and $f_\mathrm{yr} \equiv 1/(1\,\mathrm{yr})$. 
The optimal statistic $\hat{A}^2$ is given by
\begin{equation}
	\hat{A}^2 = \frac{\sum_{ab} \res_a^T {C_a^{-1} \tilde{S}_{ab} C_b^{-1} \res_b}}{\sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) } \,,
\end{equation}
where $\tilde{S}_{ab}$ is the amplitude-independent cross-correlation matrix,
\begin{equation}
	\Agw^2 \tilde{S}_{ab} = S_{ab} \,.
\end{equation}
This definition of the optimal statistic ensures that  
$\langle \hat{A}^2 \rangle = \Agw^2$. 
If $\Agw=0$, the variance of the optimal statistic is
\begin{equation}
	\sigma_0 = \left[ \sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) \right]^{-1/2} \,.
\end{equation}
For a measured value of $\hat{A}^2$, 
the significance of $\hat{A}^2 \neq 0$ is given by the signal-to-noise ratio (SNR) 
\begin{equation}
	\rho = \frac{\sum_{ab} \res_a^T {C_a^{-1} \tilde{S}_{ab} C_b^{-1} \res_b}}{ \left[ \sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) \right]^{1/2}} \,.
\end{equation}

@@

In Fig.~\ref{fig:os_datasetstats} we show the optimal statistic 
for 200 different realizations of a GWB with $\Agw = 5\times10^{-15}$ 
computed using the three techniques described above. 
For the noise-marginalized analysis, we plot 
the mean values of $\hat{A}^2$ and $\rho$. 
Using the noise values from individual noise analyses 
systematically underestimates the strength of the GWB, 
while using the noise values from a common noise analysis 
more accurately recovers the injected value. 
The fixed-noise analysis using the individual noise results finds 
$\hat{A}^2 = (7.9 \pm 6.8) \times10^{-30}$ and $\rho = 2.3 \pm 1.5$, 
averaging over realizations of the GWB. 
The fixed-noise and noise-marginalized analyses 
using the common noise results both give 
$\hat{A}^2 = (2.4\pm1.2)\times10^{-29}$ and $\rho = 4.1 \pm 1.7$.

@@

\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_A5e-15.pdf}
	\caption{Optimal statistic and SNR for 200 simulated data sets 
			containing a GWB with $\Agw = 5\times10^{-15}$. 
			The fixed-noise analysis using the individual noise values (blue) 
			systematically underestimates $\hat{A}^2$, while both the 
			fixed-noise analysis using the common noise values (orange) 
			and the noise-marginalized analysis (green) 
			accurately recover $\Agw$.}
	\label{fig:os_datasetstats}
\end{figure}

@@

The fixed-noise and noise-marginalized analyses 
using the common noise results give the same results 
for $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing data sets 
containing smaller injected values of $\Agw$. 
In Fig.~\ref{fig:os_compare} we show a P--P plot 
of the cumulative fraction of simulations for which the injected $\Agw^2$ lies within 
a given confidence interval of the measured $\hat{A}^2$. 
The confidence interval of $\hat{A}^2$ is determined 
assuming $\hat{A}^2$ follows a Gaussian distribution, 
with mean and variance $\sigma^2_{\hat{A}^2}$ 
taken from the distribution for $\hat{A}^2$ 
found from our 200 realizations of the GWB (i.e.,~the top panel of Fig.~\ref{fig:os_datasetstats}).
If $\hat{A}^2$ has a Gaussian distribution centered around $\Agw^2$, 
the curves should lie along a straight line with slope equal to unity (the dotted, diagonal lines in Fig.~\ref{fig:os_compare}).

@@

Our ability to distinguish between different spatial correlations 
depends on the strength of the GWB 
and the angular separations between pulsar pairs, $\theta_{ab}$. 
We can determine the overlap between ORFs corresponding to different spatial correlations 
by computing the ``match statistic'' \citep{cs2016},
\begin{equation}
	\bar{M} = \frac{\sum_{a,b \neq a} \Gamma_{ab} \Gamma'_{ab}}{\sqrt{ \left( \sum_{a, b \neq a} \Gamma_{ab} \Gamma_{ab} \right) \left( \sum_{a, b \neq a} \Gamma'_{ab} \Gamma'_{ab} \right)}} \,,
\end{equation}
where $\Gamma$ and $\Gamma'$ are two different ORFs. 
For the 18 pulsars used in these simulations, the 
match statistic for monopole and Hellings-Downs correlations is $\bar{M} = 0.264$, 
and the match statistic for dipole and Hellings-Downs correlations is $\bar{M} = 0.337$. 
These match statistics describe a fundamental limit on our ability 
to identify the spatial correlations of a common red signal as Hellings-Downs 
rather than monopole or dipole 
that depends only on the number of pulsars in our PTA and their sky positions.

Figure~\ref{fig:os_ORF} shows the noise-marginalized 
mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 200 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
Using quadrupole spatial correlations gives $\hat{A}^2 = (2.4 \pm 1.2) \times 10^{-29}$, 
while using monopole spatial correlations gives $\hat{A}^2 = (2.5 \pm 3.1) \times 10^{-30}$, 
and dipole spatial correlations gives $\hat{A}^2 = (5.3 \pm 4.2) \times 10^{-30}$. 
We find a noise-marginalized mean SNR above 1.0 in 97\% of our simulated data sets 
using the quadrupole ORF, and in 50\% and 68\% of our simulated data sets 
using the monopole and dipole ORFs, respectively. 
The mean SNR using the quadrupole ORF, averaged over realizations of the GWB, is 4.1, 
and we find an SNR greater than this using the monopole and dipole ORFs in just 
3\% and 3.5\% of our simulations, respectively.
\begin{figure}[t]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_spatial_A5e-15.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 200 simulated data sets 
			containing an injected GWB with $\Agw = 5\times10^{-15}$. 
			We compare the values of the mean optimal statistic and the SNR found 
			using monopole (blue), dipole (orange), and quadrupole (green) spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

This overlap between the monopole, dipole, and quadrupole ORFs also means that 
a common red process that does not have Hellings-Downs correlations  
may be confused for a GWB. Figure~\ref{fig:os_dipole} shows the results of 25 simulations 
containing a stochastic signal with dipole spatial correlations. 
Although a dipole signal has been injected, 
the Hellings-Downs ORF gives a mean SNR greater than 5 in 84\% of the simulations. 
However, both the monopole and Hellings-Downs ORFs give smaller values of 
the mean $\hat{A}^2$ and mean SNR compared to the dipole ORF. 
This demonstrates the importance of comparing the SNR from different spatial correlations 
when determining the type of spatial correlations present.
\begin{figure}[t]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_dipole_A5e-15.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 25 simulated data sets 
			containing an injected stochastic signal with dipole spatial correlations and $A = 5\times10^{-15}$. 
			We compare the values of the optimal statistic and mean SNR 
			found using monopole (blue), dipole (orange), 
			and quadrupole (green) spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_dipole}
\end{figure}

@@

We generated 725 scrambled ORFs 
using a Monte Carlo algorithm. 
We required the scrambled ORFs to have $\bar{M} < 0.2$ 
with respect to the true ORF and each other. 
This threshold was chosen to be comparable to the match statistics 
between the Hellings-Downs ORF with monopole and dipole ORFs 
given in Sec.~\ref{sec:spatial}. 
We did not chose a smaller threshold because significantly more time 
would have been needed to generate 725 scrambled ORFs. 
For each simulation, we computed the 
noise-marginalized mean optimal statistic and mean SNR 
for each scrambled ORF, 
and compared them to the values found using the true ORF.

Figure~\ref{fig:skyscrambles_dataset_sample} shows the results of a sky scramble analysis 
for a sample data set with $\Agw=5\times10^{-15}$. 
For this particular realization of the GWB, 
none of the 725 scrambled ORFs resulted in a mean SNR greater than 
the mean SNR using the true ORF ($p < 0.0014$).
In Fig.~\ref{fig:skyscrambles_pvalues}, we plot the distribution of $p$-values of the 725 sky scrambles. 
For a GWB with $\Agw = 5\times10^{-15}$, 95\% of the simulations have $p \leq 0.05$, 
and 74\% of the simulations have $p \leq 0.003$. 
For a GWB with $\Agw = 10^{-15}$, 76\% of the simulations have $p \leq 0.05$, 
and 39\% have $p \leq 0.003$. 
This shows that for smaller values of \Agw, 
there is a greater chance that noise fluctuations will appear 
to have the spatial correlations of the GWB.
\begin{figure}[t]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_scrambled_dataset11.pdf}
	\caption{Comparison between the noise-marginalized mean optimal statistic and mean SNR 
			with and without sky scrambles for a simulated data set 
			containing a GWB with $\Agw = 5\times10^{-15}$. 
			None of the 725 scrambled ORFs gave a mean SNR as large 
			as the mean SNR using the true ORF ($p < 0.0014$).}
	\label{fig:skyscrambles_dataset_sample}
\end{figure}

\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_scrambled.pdf}
	\caption{Distribution of $p$-values for the noise-marginalized optimal statistic mean SNR 
			using the true ORF compared to 725 sky scrambles. 
			We show results for simulations with $\Agw=5\times10^{-15}$ (blue) and $\Agw=10^{-15}$ (orange).}
	\label{fig:skyscrambles_pvalues}
\end{figure}

@@

The definitive signature of a GWB in PTA data is spatial correlations 
described by the Hellings-Downs curve. 
Searching for these using a full Bayesian approach is computationally expensive, 
requiring many weeks on a supercomputing cluster. 
In contrast, the optimal statistic can be computed in seconds. 
In this paper, we introduce an improved method 
for computing the optimal statistic, 
which uses the output from a Bayesian analysis 
for individual and common red signals 
to marginalize the optimal statistic over the individual pulsars' red noises. 
As shown in Sec.~\ref{sec:marg_os}, 
the noise-marginalized optimal statistic more accurately recovers the GWB amplitude 
than the fixed-noise optimal statistic, which underestimates the GWB amplitude 
when significant red noise is present in some pulsars.

Although the noise-marginalized optimal statistic 
requires computing the optimal statistic thousands of times, 
it is still many orders of magnitude faster than a Bayesian search. 
Furthermore, the results from a single Bayesian analysis, 
which are needed to marginalize over the red noise parameters, 
can be used to compute the optimal statistic for many different spatial correlations. 
In Sec.~\ref{sec:spatial} we use the noise-marginalized optimal statistic 
to compare the strength of monopole, dipole, and quadrupole correlations 
in simulated PTA data with a GWB. 
In Sec.~\ref{sec:skyscrambles} we use the noise-marginalized optimal statistic 
to perform sky scramble analyses, where we compare the mean SNR 
computed using the true ORF to the mean SNR computed using scrambled ORFs 
and measure the significance of HD spatial correlations through the $p$-value.

>>>>

\begin{abstract}
Observations have revealed that nearly all galaxies 
contain supermassive black holes (SMBHs) at their centers. 
When galaxies merge, these SMBHs form SMBH binaries (SMBHBs), 
which emit low-frequency gravitational waves (GWs). 
The incoherent superposition of these sources produce a stochastic GW background (GWB) 
that can be observed by pulsar timing arrays (PTAs). 
The optimal statistic is a frequentist estimator of the amplitude of the GWB 
that specifically looks for the spatial correlations between pulsars induced by the GWB. 
In this paper, we introduce an improved method for computing the optimal statistic 
that marginalizes over the red noise in individual pulsars. 
We use simulations to demonstrate that this method 
more accurately determines the strength of the GWB, 
and we use the noise-marginalized optimal statistic 
to compare the significance of monopole, dipole, and Hellings-Downs (HD) spatial correlations 
and perform sky scrambles.
\end{abstract}

@@

However, when pulsars have significant red noise, 
the optimal statistic gives biased results 
due to the strong covariance 
between the individual red noise parameters and the GWB amplitude. 
Many individual pulsars show evidence for red noise \citep{lam+2017,abb+17}, 
and uncertainty in the position of the Solar System barycenter (SSB) leads to 
a common red process in all pulsars \citep{abb+17b}. 
%In this paper, we present a technique for improving the accuracy of the optimal statistic 
%by marginalizing the optimal statistic over the individual red noise parameters 
%using the posterior distributions from a full Bayesian analysis. 
%This hybrid approach produces a more accurate 
%estimate of \Agw\ and its uncertainty. 
Here we present a technique for improving the accuracy of the optimal statistic by including an additional step: 
marginalizing over the individual pulsar red noise parameters 
using the posterior distributions from a full Bayesian analysis of all the pulsars. 
This hybrid approach produces a more precise estimate of the GWB amplitude \Agw\ 
and its uncertainty, while requiring only a few minutes more than the more traditional method of computation. 
%This method is more computationally intensive than previous methods 
%for computing the optimal statistic, but still requires only minutes. 
%Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
%can be used to compute the optimal statistic with any choice of the ORF, 
%allowing additional spatial correlations to be tested very quickly.
Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
can be used to compute the optimal statistic for any choice of spatial correlations 
simply by changing the overlap-reduction function (ORF). 
For example, clock errors lead to a common red signal with monopole spatial correlations \citep{hcm+2012}, 
while uncertainty in the SSB produces dipole spatial correlations \citep{chm+2010}. 
This technique is used to perform the frequentist searches 
for common red signals with HD, 
monopole, and dipole spatial correlations 
in the NANOGrav 11-year data set \citep{abb+17b}.

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic. 
We use simulations to compare the noise-marginalized optimal statistic 
to the optimal statistic computed with fixed noise. 
In Sec.~\ref{sec:spatial} we determine how well 
the noise-marginalized optimal statistic can 
differentiate between monopole, dipole, and HD spatial correlations. 
In Sec.~\ref{sec:skyscrambles} we use the noise-marginalized optimal statistic 
to perform sky scrambles, 
which assess the significance of HD spatial correlations 
by scrambling the pulsars' sky positions \citep{cs2016, tlb+2017}. 
We summarize our results in Sec.~\ref{sec:conclusion} 
as well as discuss future applications of the noise-marginalized optimal statistic.

@@

The optimal statistic is constructed from the autocovariance and cross-covariance matrices 
$C_a$ and $S_{ab}$, 
\begin{eqnarray}
	C_a &=& \left\langle \res_a \res_a^T \right\rangle \,, \\
	S_{ab} &=& \left. \left\langle \res_a \res_b^T \right\rangle \right|_{a \neq b} \,,
\end{eqnarray}
where $\res_a$ is a vector of the residuals of the $a$th pulsar in the PTA. 
For the GWB with power spectral density (PSD) $P_\mathrm{gw}(f)$ 
and overlap reduction function (ORF) $\Gamma_{ab}$, 
the cross-covariance matrices are
\begin{equation}
	S_{ab} = F_a \, \phi_{ab}^\mathrm{gw} \, F_b^T \,,
\end{equation}
where
\begin{equation}
	\phi_{ab}^\mathrm{gw} = \Gamma_{ab} \, P_\mathrm{gw}(f) \,,
	\label{eq:phi_gw}
\end{equation}
and the ORF is the HD curve \citep{hd1983}, 
\begin{eqnarray}
	\Gamma_{ab} &=& \frac{1}{2} \left[ 1 - \frac{1}{2} \left( \frac{1-\cos\theta_{ab}}{2} \right) \right. \nonumber \\
		&& \left. + 3 \left( \frac{1-\cos\theta_{ab}}{2} \right) \ln \left( \frac{1-\cos\theta_{ab}}{2} \right) \right]  \,,
\end{eqnarray}
where $\theta_{ab}$ is the angle between the pulsars. 
We model the PSD of the GWB as a power law:
\begin{equation}
	P_\mathrm{gw}(f) = \frac{\Agw^2}{12\pi^2} \left( \frac{f}{f_\mathrm{yr}} \right)^{-\gamma} \,,
\end{equation}
where $\gamma = 13/3$ assuming SMBHBs evolve solely due to GW emission and $f_\mathrm{yr} \equiv 1/(1\,\mathrm{yr})$. 
The optimal statistic $\hat{A}^2$ is given by
\begin{equation}
	\hat{A}^2 = \frac{\sum_{ab} \res_a^T {C_a^{-1} \tilde{S}_{ab} C_b^{-1} \res_b}}{\sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) } \,,
\end{equation}
where $\tilde{S}_{ab}$ is the amplitude-independent cross-correlation matrix,
\begin{equation}
	\Agw^2 \tilde{S}_{ab} = S_{ab} \,.
\end{equation}
This definition of the optimal statistic ensures that  
$\langle \hat{A}^2 \rangle = \Agw^2$. 
If $\Agw=0$, the variance of the optimal statistic is
\begin{equation}
	\sigma_0 = \left[ \sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) \right]^{-1/2} \,.
\end{equation}
For a measured value of $\hat{A}^2$, 
the significance of $\hat{A}^2 \neq 0$ is given by the signal-to-noise ratio (SNR) 
\begin{equation}
	\rho = \frac{\sum_{ab} \res_a^T {C_a^{-1} \tilde{S}_{ab} C_b^{-1} \res_b}}{ \left[ \sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) \right]^{1/2}} \,.
\end{equation}

@@

In Fig.~\ref{fig:os_datasetstats} we show the optimal statistic 
for 300 different realizations of a GWB with $\Agw = 5\times10^{-15}$ 
computed using the three techniques described above. 
For the noise-marginalized analysis, we plot 
the mean values of $\hat{A}^2$ and $\rho$. 
Using the noise values from individual noise analyses 
systematically underestimates the strength of the GWB, 
while using the noise values from a common noise analysis 
more accurately recovers the injected value. 
The fixed-noise analysis using the individual noise results finds 
$\hat{A}^2 = (7.9 \pm 6.8) \times10^{-30}$ and $\rho = 2.3 \pm 1.5$, 
averaging over realizations of the GWB. 
The fixed-noise and noise-marginalized analyses 
using the common noise results both give 
$\hat{A}^2 = (2.4\pm1.2)\times10^{-29}$ and $\rho = 4.1 \pm 1.7$.

@@

\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_A5e-15.pdf}
	\caption{Optimal statistic and SNR for 300 simulated data sets 
			containing a GWB with $\Agw = 5\times10^{-15}$. 
			The fixed-noise analysis using the individual noise values (blue) 
			systematically underestimates $\hat{A}^2$, while both the 
			fixed-noise analysis using the common noise values (orange) 
			and the noise-marginalized analysis (green) 
			accurately recover $\Agw$.}
	\label{fig:os_datasetstats}
\end{figure}

@@

The fixed-noise and noise-marginalized analyses 
using the common noise results give the same results 
for $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing data sets 
containing smaller injected values of $\Agw$. 
In Fig.~\ref{fig:os_compare} we show a P--P plot 
of the cumulative fraction of simulations for which the injected $\Agw^2$ lies within 
a given confidence interval of the measured $\hat{A}^2$. 
The confidence interval of $\hat{A}^2$ is determined 
assuming $\hat{A}^2$ follows a Gaussian distribution, 
with mean and variance $\sigma^2_{\hat{A}^2}$ 
taken from the distribution for $\hat{A}^2$ 
found from our 300 realizations of the GWB (i.e.,~the top panel of Fig.~\ref{fig:os_datasetstats}).
If $\hat{A}^2$ has a Gaussian distribution centered around $\Agw^2$, 
the curves should lie along a straight line with slope equal to unity (the dotted, diagonal lines in Fig.~\ref{fig:os_compare}).

@@

Our ability to distinguish between different spatial correlations 
depends on the strength of the GWB 
and the angular separations between pulsar pairs, $\theta_{ab}$. 
We can determine the overlap between ORFs corresponding to different spatial correlations 
by computing the ``match statistic'' \citep{cs2016},
\begin{equation}
	\bar{M} = \frac{\sum_{a,b \neq a} \Gamma_{ab} \Gamma'_{ab}}{\sqrt{ \left( \sum_{a, b \neq a} \Gamma_{ab} \Gamma_{ab} \right) \left( \sum_{a, b \neq a} \Gamma'_{ab} \Gamma'_{ab} \right)}} \,,
\end{equation}
where $\Gamma$ and $\Gamma'$ are two different ORFs. 
For the 18 pulsars used in these simulations, the 
match statistic for monopole and HD correlations is $\bar{M} = 0.264$, 
and the match statistic for dipole and HD correlations is $\bar{M} = 0.337$. 
These match statistics describe a fundamental limit on our ability 
to identify the spatial correlations of a common red signal as HD 
rather than monopole or dipole 
that depends only on the number of pulsars in our PTA and their sky positions.

Figure~\ref{fig:os_ORF} shows the noise-marginalized 
mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and HD spatial correlations 
for 300 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
HD ORF. 
Using HD spatial correlations gives $\hat{A}^2 = (2.4 \pm 1.2) \times 10^{-29}$, 
while using monopole spatial correlations gives $\hat{A}^2 = (2.5 \pm 3.1) \times 10^{-30}$, 
and dipole spatial correlations gives $\hat{A}^2 = (5.3 \pm 4.2) \times 10^{-30}$. 
We find a noise-marginalized mean SNR above 1.0 in 97\% of our simulated data sets 
using the HD ORF, and in 50\% and 68\% of our simulated data sets 
using the monopole and dipole ORFs, respectively. 
The mean SNR using the HD ORF, averaged over realizations of the GWB, is 4.1, 
and we find an SNR greater than this using the monopole and dipole ORFs in just 
3\% and 3.5\% of our simulations, respectively.
\begin{figure}[t]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_spatial_A5e-15.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 300 simulated data sets 
			containing an injected GWB with $\Agw = 5\times10^{-15}$. 
			We compare the values of the mean optimal statistic and the SNR found 
			using monopole (blue), dipole (orange), and HD (green) spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

This overlap between the monopole, dipole, and HD ORFs also means that 
a common red process that does not have HD correlations  
may be confused for a GWB. Figure~\ref{fig:os_dipole} shows the results of 25 simulations 
containing a stochastic signal with dipole spatial correlations. 
Although a dipole signal has been injected, 
the HD ORF gives a mean SNR greater than 5 in 84\% of the simulations. 
However, both the monopole and HD ORFs give smaller values of 
the mean $\hat{A}^2$ and mean SNR compared to the dipole ORF. 
This demonstrates the importance of comparing the SNR from different spatial correlations 
when determining the type of spatial correlations present.
\begin{figure}[t]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_dipole_A5e-15.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 25 simulated data sets 
			containing an injected stochastic signal with dipole spatial correlations and $A = 5\times10^{-15}$. 
			We compare the values of the optimal statistic and mean SNR 
			found using monopole (blue), dipole (orange), 
			and HD (green) spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_dipole}
\end{figure}

@@

We generated 725 scrambled ORFs 
using a Monte Carlo algorithm. 
We required the scrambled ORFs to have $\bar{M} < 0.2$ 
with respect to the true ORF and each other. 
This threshold was chosen to be comparable to the match statistics 
between the HD ORF with monopole and dipole ORFs 
given in Sec.~\ref{sec:spatial}. 
We did not chose a smaller threshold because significantly more time 
would have been needed to generate 725 scrambled ORFs. 
For each simulation, we computed the 
noise-marginalized mean optimal statistic and mean SNR 
for each scrambled ORF, 
and compared them to the values found using the true ORF.

Figure~\ref{fig:skyscrambles_dataset_sample} shows the results of a sky scramble analysis 
for a sample data set with $\Agw=5\times10^{-15}$. 
For this particular realization of the GWB, 
none of the 725 scrambled ORFs resulted in a mean SNR greater than 
the mean SNR using the true ORF ($p < 0.0014$).
In Fig.~\ref{fig:skyscrambles_pvalues}, we plot the distribution of $p$-values of the 725 sky scrambles 
for 300 realizations of the GWB. 
For a GWB with $\Agw = 5\times10^{-15}$, 95\% of the simulations have $p \leq 0.05$, 
and 74\% of the simulations have $p \leq 0.003$. 
For a GWB with $\Agw = 10^{-15}$, 76\% of the simulations have $p \leq 0.05$, 
and 39\% have $p \leq 0.003$. 
This shows that for smaller values of \Agw, 
there is a greater chance that noise fluctuations will appear 
to have the spatial correlations of the GWB.
\begin{figure}[t]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_scrambled_dataset11.pdf}
	\caption{Comparison between the noise-marginalized mean optimal statistic and mean SNR 
			with and without sky scrambles for a simulated data set 
			containing a GWB with $\Agw = 5\times10^{-15}$. 
			None of the 725 scrambled ORFs gave a mean SNR as large 
			as the mean SNR using the true ORF ($p < 0.0014$).}
	\label{fig:skyscrambles_dataset_sample}
\end{figure}

\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_scrambled.pdf}
	\caption{Distribution of $p$-values for the noise-marginalized optimal statistic mean SNR 
			using the true ORF compared to 725 sky scrambles from 300 realizations of the GWB. 
			We show results for simulations with $\Agw=5\times10^{-15}$ (blue) and $\Agw=10^{-15}$ (orange).}
	\label{fig:skyscrambles_pvalues}
\end{figure}

@@

The definitive signature of a GWB in PTA data is spatial correlations 
described by the HD curve. 
Searching for these using a full Bayesian approach is computationally expensive, 
requiring many weeks on a supercomputing cluster. 
In contrast, the optimal statistic can be computed in seconds. 
In this paper, we introduce an improved method 
for computing the optimal statistic, 
which uses the output from a Bayesian analysis 
for individual and common red signals 
to marginalize the optimal statistic over the individual pulsars' red noises. 
As shown in Sec.~\ref{sec:marg_os}, 
the noise-marginalized optimal statistic more accurately recovers the GWB amplitude 
than the fixed-noise optimal statistic, which underestimates the GWB amplitude 
when significant red noise is present in some pulsars.

Although the noise-marginalized optimal statistic 
requires computing the optimal statistic thousands of times, 
it is still many orders of magnitude faster than a Bayesian search. 
Furthermore, the results from a single Bayesian analysis, 
which are needed to marginalize over the red noise parameters, 
can be used to compute the optimal statistic for many different spatial correlations. 
In Sec.~\ref{sec:spatial} we use the noise-marginalized optimal statistic 
to compare the strength of monopole, dipole, and HD correlations 
in simulated PTA data with a GWB. 
In Sec.~\ref{sec:skyscrambles} we use the noise-marginalized optimal statistic 
to perform sky scramble analyses, where we compare the mean SNR 
computed using the true ORF to the mean SNR computed using scrambled ORFs 
and measure the significance of HD spatial correlations through the $p$-value.
